Scaffolder - Software for Reproducible Genome Scaffolding.

### 2011-03-08 src/article.tex
Revise results and discussion

The Scaffolder software allows computational joining of nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in the YAML syntax specify how these sequences should be
joined and the scaffolder software is used to generate the scaffold sequence.
In addition to specifying which contigs are required, the scaffold file allows
the contigs to be edited into smaller sub-sequences or reverse complemented if
necessary. The process of genome finishing may involve producing additional
oligonucleotide sequences to fill unknown regions in the scaffold and The
scaffold file format allows these additional insert sequences to bridge gaps.
These inserts can also be treated in the same manner as larger contig
sequences: trimmed and/or reverse complemented to match the corresponding gap
region. \pb

@@

The scaffold file specifies how sequence regions are joined together; however
the sequences themselves are not stored in the scaffold file and are instead
maintained as a separate FASTA file. Nucleotide sequences in a FASTA-format
file are the standard output for a genome assembler, and these may be contigs,
sets of contigs scaffolded into larger sequences, or both. These nucleotide
sequences stored in the FASTA file are referenced in the scaffold using the
first word from in the FASTA header of the corresponding sequence. Maintaining
the nucleotide sequences in a separate file preserves the unedited sequence and
decouples the data from the scaffold file describing how it should be used to
in the genome sequence. \pb

@@

The scaffold file is written using the YAML syntax and an example is shown in
Figure 1. This file illustrates the text attributes used to describe a scaffold
and how the sequences are joined together in the genome build. The basic layout
of the scaffold file is a list of entries, where each entry corresponds to
a region of sequence in the generated scaffold super-sequence. \pb

@@

The first line of the scaffold file begins with three dashes to indicate the
start of a YAML-formatted document. The first entry (highlighted in green)
begins with a dash character `-' to denote an entry in the YAML list. This is
a requirement of the YAML format: that each entry begins with a dash line. The
next line is indented by two spaces as whitespace is required to group similar
attributes together. The ``sequence'' tag indicates that this entry corresponds
to a sequence in the FASTA file and the following line indicates the name of
this sequence using the ``source'' tag. As described the first word of the
FASTA header is used to identify which sequence is selected from the file.
Together these three lines describe the first entry in the scaffold as a simple
sequence using the sequence named `sequence1'. On the right hand side of Figure
1 this produces the first region in the scaffold, also shown in green. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the sequence named
`sequence2' to the scaffold. This entry demonstrates how this sequence may be
manipulated prior to addition to the scaffold. The `start' and `stop' tags trim
the sequence to these coordinates inclusively, while the ``reverse'' tag
instructs scaffolder to also reverse complement the sequence. In the scaffolded
genome shown in Figure 1 this completes the scaffold sequence. \pb

The final entry in the scaffold also includes the ``inserts'' tag to add
additional regions of sequence. These inserts are also added as a YAML list,
with each insert beginning with a dash. The first insert, shown in purple,
similar attributes to that of a sequence entry; the reverse, start and stop
tags are used to trim and reverse complement the insert. Similarly the `source'
tag identifies the corresponding FASTA sequence as `insert1'. The ``open'' and
``close'' tags are specific to inserts and determine where the insert is added
in the enclosing sequence. The region of the sequence inside these coordinates
is inclusively replaced by the specified insert sequence. This is visualised in
the putative genome in Figure 1 by the black lines bisecting the blue sequence.
\pb

@@

Scaffolder provides a standardised set of Ruby classes and methods (API) for
interacting with the scaffold. This allows Scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as Rake. In
addition Scaffolder provides a command line interface (CLI) to validate the
scaffold file and build a draft super sequence. The Scaffolder CLI behaves as
a standard Unix tool and returns appropriate exit codes and manual pages. The
use of both these Scaffolder interfaces is outlined in detail on the scaffolder
website (\scaffolder). \pb

@@

Scaffolding requires joining assembled contigs and insert additional sequence
data using a combination of computational and laboratory methods. The process
of manually editing a scaffold is inherently hard to reproduce and introduces
the possibility of irreducible techniques and/or human error. The aims of
the Scaffolder software are twofold: 1) to provide software that is easy to
install and simplifies the task of genome finishing; and 2) to facilitate
reproducibility when joining large nucleotide sequences together to create
a genome sequence. \pb

Scaffolder was designed to be as simple to use and, assuming the Ruby and
RubyGems software are present, Scaffolder can be installed in a single command
line step. This should negate any possible barriers to entry which require
manual compilation of source code. Scaffolder also uses a minimal and compact
syntax to describe the genome scaffold. This syntax is simple to construct and
edit whilst being succinct and readable. The goal of this syntax is to make
genome scaffolds easy to make in a common text editor; however the use of word
processors for this task is discouraged. \pb

Constructing a genome using a text file makes generating a scaffold super
sequence both reproducible and deterministic for the same scaffold and FASTA
sequence file. In comparison, joining large nucleotide sequences by hand cannot
be reliably reproduced. The scaffold file furthermore provides a human readable
description of how the scaffold is constructed. Configuring the final sequence
in the scaffold file means the build is easier to edit once already
constructed. The YAML text format allows comparison of differences between
scaffold builds using standard Unix tools such as diff. This makes scaffold
files easier to store in version control systems and allows genome finishers to
use methods similar to those in software development. \pb

>>>>

The Scaffolder software facilitates reproducibly joining nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in the YAML syntax specify how these sequences should
be joined and the scaffolder software is used to generate the scaffold
sequence from these instructions. In addition to specifying which contigs are
required, the scaffold file allows the contigs to be edited into smaller
sub-sequences or reverse complemented if necessary. The process of genome
finishing may involve producing additional oligonucleotide sequences to fill
unknown regions in the scaffold and the Scaffolder allows these additional
insert sequences to be used to fill these gaps. These inserts can also be
treated in the same manner as larger contig sequences: trimmed and/or reverse
complemented to match the corresponding gap region size and orientation. \pb

@@

The scaffold file specifies how sequence regions are joined together; however
the sequences themselves are not stored in the scaffold file and are instead
maintained as a separate FASTA file. Nucleotide sequences in a FASTA-format
file are the standard output for a genome assembler, and these may be contigs,
sets of contigs scaffolded into larger sequences, or both. These nucleotide
sequences stored in the FASTA file are referenced in the scaffold using the
first word from in the FASTA header of the corresponding sequence. Maintaining
the nucleotide sequences in a separate file preserves the unedited sequence
and decouples the data from the specification of how it should be used to
produce the genome sequence. \pb

@@

The scaffold file is written using the YAML syntax and an example is shown in
Figure 1. This file illustrates the text attributes used to describe
a scaffold and how the sequences are correspondingly joined together in the
genome build. The basic layout of the scaffold file is a list of entries,
where each entry corresponds to a region of sequence in the generated scaffold
super-sequence. \pb

@@

The first line of the scaffold file in Figure 1 begins with three dashes to
indicate the start of a YAML-formatted document. The first entry (highlighted
in green) begins with a dash character `-' to denote an entry in the YAML
list. This is a requirement of the YAML format: that each entry begins with
a dash line. The next line is indented by two spaces where whitespace is
required to group similar attributes together. The ``sequence'' tag indicates
that this entry corresponds to a sequence in the FASTA file and the following
line indicates the name of this sequence using the ``source'' tag. The first
word of the FASTA header is used to identify which sequence is selected from
the file. Together these three lines describe the first entry in the scaffold
as a simple sequence using the sequence named `sequence1'. On the right hand
side of Figure
1 this produces the first region in the scaffold, also shown in green. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the sequence named
`sequence2' to the scaffold. This entry demonstrates how this sequence may be
manipulated prior to addition to the scaffold. The `start' and `stop' tags trim
the sequence to these coordinates inclusively, while the ``reverse'' tag
instructs scaffolder to also reverse complement the sequence. In the putative
scaffold shown in Figure 1 this completes the scaffold sequence. \pb

The final entry in the scaffold also includes the ``inserts'' tag to add
additional regions of sequence. These inserts are also added as a YAML list,
with each insert beginning with a dash. The first insert, shown in purple,
uses similar attributes to that of a sequence entry; the reverse, start and
stop tags are used to trim and reverse complement the insert. Similarly the
`source' tag identifies the corresponding FASTA sequence as `insert1'. The
``open'' and ``close'' tags are specific to inserts and determine where the
insert is added in the enclosing sequence. The region of the sequence inside
these coordinates is inclusively replaced by the specified insert sequence.
This is visualised in the putative scaffold in Figure 1 by the black lines
bisecting the blue sequence. \pb

@@

Scaffolder provides a standardised set of Ruby classes and methods (API) for
interacting with the scaffold. This allows Scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as Rake. In
addition Scaffolder provides a command line interface (CLI) to validate the
scaffold file and build the draft super sequence. The Scaffolder CLI behaves
as a standard Unix tool and returns appropriate exit codes and manual pages.
The use of both these Scaffolder interfaces is outlined in detail on the
scaffolder website (\scaffolder). \pb

@@

Scaffolding an incomplete genome assembly requires joining contigs and
additional gap-filling sequence using a combination of computational and
laboratory methods. The process of manually editing a scaffold is inherently
hard to reproduce and introduces the possibility of irreducible techniques
and/or human error. In respect to these drawbacks the aims of the Scaffolder
software are twofold: 1) to provide software that is easy to install and
simplifies the task of genome finishing; and 2) to facilitate reproducibility
in the scaffolding and finishing stage of a genome project. \pb

Scaffolder was designed to be as simple to use and, assuming the Ruby and
RubyGems software are present, can be installed in a single command line step.
This should negate any possible barriers to entry which require manual
compilation of source code. Scaffolder also uses a minimal and compact syntax
to describe how the genome scaffold sequence should be generated. This syntax
is simple to construct and edit whilst being succinct and readable. The goal
of this syntax is to make genome scaffolds easy to write in a common text
editor. \pb

Constructing a genome using by specifying the scaffold organisation in text
file makes generating a scaffold super sequence both reproducible and
deterministic for the same file and set of FASTA sequences. In comparison,
joining large nucleotide sequences by hand cannot be reliably reproduced. The
scaffold file furthermore provides a human readable description of how the
scaffold is constructed. Configuring the final sequence in the scaffold file
means the build is easier to edit once already constructed. The YAML text
format allows comparison of differences between scaffold builds using standard
Unix tools such as diff. This makes scaffold files easier to store in version
control systems and allows genome finishers to use methods similar to those in
software development. \pb
